### 2014-03-27 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding fourier mode analysis to multilevel note

\documentclass[note]{TechNote}
\usepackage[centertags]{amsmath}
\usepackage{amssymb,amsthm}
\usepackage[mathcal]{euscript}
\usepackage{tabularx}
\usepackage{cite}
\usepackage{c++}
\usepackage{tmadd,tmath}

@@

To elucidate the effect of a given solution technique on a given error
mode in the problem, we can assign an initial guess of $\ve{x}^0$ to
be a chosen Fourier mode:
\begin{equation}
  \ve{x}^0_i = \sin\Bigg( \frac{ik\pi}{N} \Bigg)\:,
  \label{eq:fourier_mode}
\end{equation}
where $\ve{x}^0_i$ is the $i^{th}$ component of the initial guess and
k is the wave number of the chosen Fourier mode. First, we will look at
the performance Richardson's iteration as the smoother which is
equivalently a Jacobi iteration when Jacobi preconditioning is used:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I}-\ve{M}^{-1}\ve{A})\ve{x}^k\:,
\end{equation}
where $k$ is the iteration index. Figure~\ref{fig:richardson} gives
infinity norm of the error in the solution vector\footnote{The
  solution to the homogenous problem is zero and therefore
  $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$} as a function of
iterations for wave numbers of 1, 5, and 10 on a grid with N = 100.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{richardson.png}
  \end{center}
  \caption{\textbf{Convergence of Richardson's iteration as a function
      of iteration for a grid of size N = 100.} \textit{Richardson's
      iteration performs better for larger wave numbers and therefore
      more oscillatory modes.}}
  \label{fig:richardson}
\end{figure}
Immediately we note that the larger the wave number the better
Richardson's iteration performs. Per the spectral analysis in
\cite{briggs_multigrid}, this iteration sequence perfoms better for
higher wave numbers as the eigenvalue spectrum spans a smaller space
then less oscillatory modes. This behavior motivated the multigrid
approach where moving a smooth mode to a coarser grid makes that mode
appear more oscillatory relative to that coarser grid, thus improving
convergence for that particular mode.

Next we perform the same calculations using the adjoint Monte Carlo
solver presented in \cite{evans_monte_2012}. Before doing this
however, we must first modify the linear problem as the Monte Carlo
solver is a direct method and therefore a homogenous problem with a
right hand side of $\ve{0}$ will yield no samples. Instead, we will
solve the residual problem:
\begin{equation}
  \ve{A}\ve{d} = \ve{r}\:,
  \label{eq:residual_problem}
\end{equation}
where the residual of the homogenous problem is:
\begin{equation}
  \ve{r} = -\ve{A}\ve{x}^0\:,
  \label{eq:homogenous_residual}
\end{equation}
and the solution is computed as:
\begin{equation}
  \ve{x} = \ve{x}^0 + \ve{d}\:.
  \label{eq:residual_solution}
\end{equation}
Forming the problem in this way lets us directly apply the adjoint
Monte Carlo method to the homogenous problem and then apply the
effective correction, $\ve{d}$, to the initial guess to give the
solution. This approach is also equivalent to performing a single
iteration of Halton's method \cite{halton_sequential_1962}. Using this
formulation we can again solve the model problem with wave numbers of
1, 5 and 10 on a grid of size N=1 but this time we vary the number of
histories used to compute the solution instead of the number of
iterations. Figure~\ref{fig:adjoint_mc} gives the results of this
calculation.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{adjoint_mc.png}
  \end{center}
  \caption{\textbf{Convergence of the adjoint Monte Carlo method as a
      function of sampled histories for a grid of size N = 100.}
    \textit{Adjoint Monte Carlo performs better for smaller wave
      numbers and therefore smoother modes.}}
  \label{fig:adjoint_mc}
\end{figure}
Surprisingly, the Monte Carlo method performs better for smooth modes
then more oscillatory modes\footnote{This behavior was also observed
  for the forward Monte Carlo method presented in
  \cite{evans_monte_2012}}. The results of these calculations are
counterintuitive given the fact that the Monte Carlo solver is
effectively a stochastic realization of of Richardson's iteration and
therefore one should expect the same spectral behavior from the
results.

Looking at the timing results in Table~\ref{tab:mc_timing}, we see
that the behavior of the Monte Carlo solver is in fact consistent with
expectations.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cc}\hline\hline
      \multicolumn{1}{c}{\textbf{Wave Number}} & 
      \multicolumn{1}{c}{\textbf{Time per History (s)}} \\
      \hline
      1 & 1 \\
      5 & 0.85 \\
      10 & 0.83 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Normalized average CPU time per history.}}
  \label{tab:mc_timing}
\end{table}
Timing results show that the average time required to compute an
entire history in the Monte Carlo solver decreases as a function of
wave number. We showed analytically in \cite{slattery_2013} that the
length of the random walk is equivalent to the number of Richardson
iterations that would be required to achieve a given convergence
criteria. In Figure~\ref{fig:richardson}, we see that fewer iterations
are required to converge larger wave numbers and therefore we should
also expect shorter random walks in the Monte Carlo solver and
therefore a faster time to solution as observed in
Table~\ref{tab:mc_timing}.

>>>>

\documentclass[note]{TechNote}
\usepackage[centertags]{amsmath}
\usepackage{amssymb,amsthm}
\usepackage[mathcal]{euscript}
\usepackage{tabularx}
\usepackage{cite}
\usepackage{c++}
\usepackage{tmadd,tmath}
\usepackage{caption}
\usepackage{subcaption}

@@

To elucidate the effect of a given solution technique on a given error
mode in the problem, we can assign an initial guess of $\ve{x}^0$ to
be a chosen Fourier mode:
\begin{equation}
  \ve{x}^0_i = \sin\Bigg( \frac{ik\pi}{N} \Bigg)\:,
  \label{eq:fourier_mode}
\end{equation}
where $\ve{x}^0_i$ is the $i^{th}$ component of the initial guess and
k is the wave number of the chosen Fourier mode. First, we will look at
the performance Richardson's iteration as the smoother which is
equivalently a Jacobi iteration when Jacobi preconditioning is used:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I}-\ve{M}^{-1}\ve{A})\ve{x}^k\:,
\end{equation}
where $k$ is the iteration index. Figure~\ref{fig:richardson} gives
infinity norm of the error in the solution vector\footnote{The
  solution to the homogenous problem is zero and therefore
  $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$.} as a function of
iterations for wave numbers of 1, 5, and 10 on a grid with N = 100.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{richardson.png}
  \end{center}
  \caption{\textbf{Convergence of Richardson's iteration as a function
      of iteration for a grid of size N = 100.} \textit{Richardson's
      iteration performs better for larger wave numbers and therefore
      more oscillatory modes.}}
  \label{fig:richardson}
\end{figure}
Immediately we note that the larger the wave number the better
Richardson's iteration performs. Per the spectral analysis in
\cite{briggs_multigrid}, this iteration sequence perfoms better for
higher wave numbers as the eigenvalue spectrum spans a smaller space
then less oscillatory modes. This behavior motivates the multigrid
approach where moving a smooth mode to a coarser grid makes that mode
appear more oscillatory relative to that coarser grid, thus improving
convergence for that particular mode.

Next we perform the same calculations using the adjoint Monte Carlo
solver presented in \cite{evans_monte_2012}. Before doing this
however, we must first modify the linear problem as the Monte Carlo
solver is a direct method and therefore a homogenous problem with a
right hand side of $\ve{0}$ will yield no samples. Instead, we will
solve the residual problem:
\begin{equation}
  \ve{A}\ve{d} = \ve{r}\:,
  \label{eq:residual_problem}
\end{equation}
where the residual of the homogenous problem is:
\begin{equation}
  \ve{r} = -\ve{A}\ve{x}^0\:,
  \label{eq:homogenous_residual}
\end{equation}
and the solution is computed as:
\begin{equation}
  \ve{x} = \ve{x}^0 + \ve{d}\:.
  \label{eq:residual_solution}
\end{equation}
Forming the problem in this way lets us directly apply the adjoint
Monte Carlo method to the homogenous problem and then apply the
effective correction, $\ve{d}$, to the initial guess to give the
solution. This approach is also equivalent to performing a single
iteration of Halton's method \cite{halton_sequential_1962}. Using this
formulation we can again solve the model problem with wave numbers of
1, 5 and 10 on a grid of size N=1 but this time we vary the number of
histories used to compute the solution instead of the number of
iterations. Figure~\ref{fig:adjoint_mc} gives the results of this
calculation.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{adjoint_mc.png}
  \end{center}
  \caption{\textbf{Convergence of the adjoint Monte Carlo method as a
      function of sampled histories for a grid of size N = 100.}
    \textit{Adjoint Monte Carlo performs better for smaller wave
      numbers and therefore smoother modes.}}
  \label{fig:adjoint_mc}
\end{figure}
Surprisingly, the Monte Carlo method performs better for smooth modes
then more oscillatory modes\footnote{This behavior was also observed
  for the forward Monte Carlo method presented in
  \cite{evans_monte_2012}.}. The results of these calculations are
counterintuitive given the fact that the Monte Carlo solver is
effectively a stochastic realization of of Richardson's iteration and
therefore one should expect the same spectral behavior from the
results.

Looking at the timing results in Table~\ref{tab:mc_timing}, we see
that the behavior of the Monte Carlo solver is in fact consistent with
expectations.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cc}\hline\hline
      \multicolumn{1}{c}{\textbf{Wave Number}} & 
      \multicolumn{1}{c}{\textbf{Time per History (s)}} \\
      \hline
      1 & 1 \\
      5 & 0.85 \\
      10 & 0.83 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Normalized average CPU time per history.}}
  \label{tab:mc_timing}
\end{table}
Timing results show that the average time required to compute an
entire history in the Monte Carlo solver decreases as a function of
wave number. We showed analytically in \cite{slattery_2013} that the
length of the random walk is equivalent to the number of Richardson
iterations that would be required to achieve a given convergence
criteria. In Figure~\ref{fig:richardson}, we see that fewer iterations
are required to converge larger wave numbers and therefore we should
also expect shorter random walks in the Monte Carlo solver and
therefore a faster time to solution as observed in
Table~\ref{tab:mc_timing}. However, this does not indicate why the
Monte Carlo solvers perform better for smoother modes rather than
oscillatory modes. Fortunately, the explanation for this behavior is
simple. Consider the plots for the Fourier modes with $k = 1$ and $k =
10$ given in Figure~\ref{fig:fourier_modes}.
\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.5\textwidth}
    \includegraphics[width=\textwidth]{mode_1.pdf}
    \caption{\textbf{Fourier mode with $k = 1$.}}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \includegraphics[width=\textwidth]{mode_10.pdf}
    \caption{\textbf{Fourier mode with $k = 10$.}}
  \end{subfigure}
  \caption{\textit{The more oscillatory the mode the finer the
      structure of the error. Finer structures require more samples to
      resolve them.}}
  \label{fig:fourier_modes}
\end{figure}
The higher the wave number is the more oscillatory the mode
appears. To resolve these fine structures in the error we therefore
naturally require more Monte Carlo samples to distinguish them from
one another.

Based on this, it may not seem that applying Monte Carlo in a
multigrid context will be useful considering that error modes appear
more oscillatory as the grid is coarsened. This would indicate that in
order to achieve the same amount of error on a coarse grid
representation of the problem, one would be required to compute many
more samples to resolve the resulting finer structures. However, we
may gain a computational advantage for two reasons. First, consider
the timing results in Table~\ref{tab:mc_timing}. As the error
structure becomes more oscillatory, the time it takes to compute a
sample decreases. Second, if we coarsen the grid the time it takes to
compute a sample will also decrease due to the decreased problem
size. Therefore, it is plausible that we might observe an improvement
in run times for the Monte Carlo problem by using a multigrid
approach. This then motivates applying recently developed techniques
in multilevel Markov chain Monte Carlo to the adjoint solver as a
potential means of accelerating time to solution.
